---
title: "**Homework 2**"
author: "Shaliha Guarnera - SDS 315 - scg2595"
date: "January 2024"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
---

[https://github.com/shalihaguarnera/SDS315HW2.git](https://github.com/shalihaguarnera/SDS315HW2.git)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}

profs <- read.csv("profs.csv")
bikeshare <- read.csv("bikeshare.csv")
capmetro <- read.csv("capmetro_UT.csv")
billboard <- read.csv("billboard.csv")
library(ggplot2)
library(tidyverse)
library(RColorBrewer)
library(kableExtra)
library(dplyr)
library(rvest)
library(sas7bdat)
library(stringr)
library(lubridate)
library(reshape2)
library(htmltools)

```

\newpage

## **Problem 1: Beauty, or Not, in the Classroom**

### **Part A**

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Creating a histogram of the distribution of course evaluation scores using ggplot2
ggplot(profs) + 
  geom_histogram(aes(x=eval), col='palevioletred3', fill='lightpink') + 
  xlab('Course Evaluation Scores') + 
  ylab('Frequency') + 
  ggtitle('Distribution of Course Evaluation Scores') +
  theme_minimal()

```

The distribution of the instructor's average teaching evaluation score is left-skewed with a median of `r median(profs$eval)` and a range of `r max(profs$eval) - min(profs$eval)`. There is a gap in the distribution between the evaluation scores 2 and 3. The course evaluation score with the greatest amount of courses is 4, and there are few courses with scores between 2 and 3.

### **Part B**

```{r echo=FALSE, warning=FALSE, message=FALSE}

# creating a faceted box plot of the distribution of course evaluation scores, faceted by whether the professor is a native english speaker
ggplot(profs) + 
  geom_boxplot(aes(x=native, y=eval, fill=native), col='palevioletred3') + 
  labs(title='Distribution of Course Evaluation Scores by Native English Speaking Status', y='Course Evaluation Scores', x='Native English Speaker (Yes/No)', fill='Native English Speaker') + 
  scale_fill_manual(values=c('lightpink', 'lightblue')) +
  theme_minimal() +
  coord_flip()

```

The distribution of course evaluation scores for instructors who are native English speakers have a higher median course evaluation score compared to instructors who are not native English speakers. This could have occurred based on the ability for the students to understand the professors and their lectures, thus affecting their understanding of the course material. The native English-speaking instructors have a greater range and were left-skewed, while the non-native English-speaking professors have more of a bell-shaped distribution.

### **Part C**

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Creating a faceted histogram of the distribution of course evaluation scores faceted by professor gender
ggplot(profs) + 
  geom_histogram(aes(x=eval, fill=gender), col='palevioletred3') + 
  labs(title='Distribution of Course Evaluation Scores by Gender', y='Frequency', x='Course Evaluation Scores', fill='Gender') + 
  facet_wrap(~gender) + 
  theme_minimal() + 
  scale_fill_manual(values=c('lightpink', 'lightblue'))

```

Both distributions of course evaluation scores by gender are left-skewed and have outliers between the scores 2 and 3. Male instructor scores are more frequent in comparison to female instructor scores. Male instructors have a higher median evaluation score, as they are the more frequent gender in the dataset.

### **Part D**

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Creating a scatterplot of the relationship between professor attractiveness and course evaluation scores
ggplot(profs) + geom_point(aes(x=beauty, y=eval), col='lightpink') + 
  xlab("Professor's Physical Attractiveness") + 
  ylab('Course Evaluation Scores') + 
  ggtitle('Course Evaluation Scores and Professor Attractiveness') + 
  theme_minimal()

```

The relationship between a professor's physical attractivenes and course evaluation scores have a correlation coefficient of `r round(cor(profs$beauty, profs$eval), 2)`. This signifies a relatively weak positive correlation between the two variables. There is no sufficient evidence to suggest that instructor physical attractiveness and course evaluation scores are related with one another.

## **Problem 2: Bike Sharing**

### **Plot A**

```{r echo=FALSE, warning=FALSE, message=FALSE}

#Creating a subset grouping the bikeshare observations by hour of the day, and calculating the average total bike rentals for each hour of the day
avg_rentals <- bikeshare %>% 
  group_by(hr) %>% 
  summarize(avg_total = mean(total))

#Creating a line graph depicting the average total bike rentals by hour of the day
ggplot(avg_rentals) + geom_line(aes(x=hr, y=avg_total), col='palevioletred3') + 
  xlab('Hours of the Day') + 
  ylab('Average Hourly Bike Rentals') + 
  ggtitle('Average Hourly Bike Rentals Across All Hours of the Day') + 
  theme_minimal()

```

The line graph above illustrates the average hourly bike rentals (y-axis, measured as a count of bike rentals) throughout all hours of the day (x-axis, measured in hours). The hours of the day are signified as numbers from 0 to 23, starting with 0 as 12 AM and ending with 23 as 11 PM. The average hourly bike rentals decreases from 12 AM, then begins to increase at around 5 AM and peaks around 8 AM before it decreases again. The rentals reach a trough at 10 AM. After gradually increasing, remaining constant, and decreasing from 10 AM to 4 PM, the rentals peak again near 5 PM, in which it reaches its greatest average bike rentals. The rentals then decrease until the end of the day at 11 PM. The rentals were at its lowest in the beginning of the day near 4 AM and at its greatest in the evening. The two peaks in rentals were most likely due to the occurrence of the beginning and end of workdays, with employers and students going to an from work/school, thus needing to rent bikes for transportation. 

### **Plot B**

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Merging the subset with the mean total bike rentals by hour to the original bikeshare data set
avg_rentals <- left_join(bikeshare, avg_rentals, by='hr')
avg_rentals$workday <- ifelse(avg_rentals$workingday == 1, 'workday', 'nonworkday')
avg_rentals_new <- avg_rentals %>% 
  group_by(workday, hr) %>% 
  summarize(avg_total = mean(total))

ggplot(avg_rentals_new) + 
  geom_line(aes(x=hr, y=avg_total, col=workday)) + 
  labs(title='Average Total Bike Rentals Across Hours of the Day by Working Day', x='Hour of the Day', y='Average Total Bike Rentals', color='Day Type') + 
  scale_color_manual(values=c('workday'='lightpink', 'nonworkday'='lightblue')) + 
  facet_wrap(~workday, scales='free_y') + theme_minimal() 

```

The faceted line graph represents the average total bike rentals by hour of the day (y-axis, measured as a count of bike rentals) throughout all hours of the day (x-axis, measured in hours). The graph is similar to the previous line graph from Plot A, although it is faceted into two different graphs based on whether the day was a workday or a non-workday. The hours of the day are signified as numbers from 0 to 23, starting with 0 as 12 AM and ending with 23 as 11 PM. The average total bike rentals for both types of days both began decreasing in the early hours of the morning, and significantly increased after 5 AM and throughout the rest of the morning. However, the average bike rentals for workdays reaches troughs in the middle of the day while non-workdays did not. This can be explained by the idea that many would not be renting bikes in the middle of the day as some are at work or in school. Since non-workdays do not have work and school, more people are available to rent the bikes. Both types of days gradually decrease as the day comes to an end. 

### **Plot C**

```{r echo=FALSE, warning=FALSE, message=FALSE}

bike9am <- bikeshare %>% 
  filter(hr==9) %>% 
  group_by(weathersit, workingday) %>%
  summarize(avg_total=mean(total))

ggplot(bike9am) + 
  geom_bar(aes(x=weathersit, y=avg_total, fill=factor(workingday)), stat = 'identity', col='palevioletred3') + 
  scale_fill_manual(name='Day Type', labels=c('Non-Workday', 'Workday'), values=c('lightpink', 'lightblue'))+
  facet_wrap(~workingday, labeller=labeller(workingday=c('0'='Non-Workday', '1'='Workday'))) +
  labs(title='Average Ridership During 9 AM by Weather Situation Code and Day Type', x='Weather Situation Code', y='Average Total Ridership') +
  theme(legend.position='none') +
  theme_minimal()

```

The given bar plot depicts average total bike rentals by weather situation code (1-3) and type of day (workday or non-workday), faceted by day type. The x-axis contains the weather situation code, with code 1 consisting of clear skies, few clouds, and/or partly cloudy conditions, code 2 with cloudy skies, mist, and/or broken clouds, and code 3 with light snow, light rain, thunderstorms, and/or scattered clouds. The y-axis is the average total bike rentals, measured by the count of total bike rentals a given hour. Both workdays and non-workdays had code 3 as the weather situation code with the least average total ridership, and code 1 as the weather situation code with the greatest average total ridership, respective to the day types. Across workdays and non-workdays, more people prefer to rent bikes during good weather conditions as it makes it easier to travel. Many do not want to rent bikes with bad weather conditions, such as thunderstorms, leading to a lowered average total bike rentals.

## **Problem 3: Capital Metro UT Ridership**

### **Facet Plot 1**

```{r echo=FALSE, warning=FALSE, message=FALSE}

capmetro = mutate(capmetro,
                  day_of_week = factor(day_of_week,
                                       levels=c("Mon", "Tue", "Wed","Thu", "Fri", "Sat", "Sun")),
                  month = factor(month,
                                 levels=c("Sep", "Oct","Nov")))

avg_boardings <- capmetro %>% 
  group_by(hour_of_day, day_of_week, month) %>% 
  summarize(avg_boarding = mean(boarding))

avg_boardings <- merge(avg_boardings, capmetro)

ggplot(avg_boardings) + geom_line(aes(x=hour_of_day, y=avg_boarding, col=factor(month))) + 
  facet_wrap(~day_of_week) +
  scale_color_manual(values=c('lightpink', 'lightblue','palevioletred3')) +
  labs(title='Average Boardings by Hour, Day of Week, and Month, Across Days', x='Hour of the Day', y= 'Average Boarding', color='Month') +
  theme_minimal()

```

The faceted grouped line graphs are of average number of boardings on any Capital Metro bus on the UT Austin campus, in a 15-minute window. The average boardings are grouped by hour of the day, day of the week, and month. The graphs are faceted into seven separate graphs for each day of the week, and with separate lines for the months September, October, and November. The x-axis is the hour of the day (in 24-hour time) and the y-axis is the average boarding, measured by the number of passengers boarding the buses. For all three of the months, Saturday and Sunday had the least amount of boardings, with little change in the average boardings, thus remaining relatively constant. The hour of peak average boardings across the days of the week remain broadly similar from Monday to Friday, excluding Saturday and Sunday. From Monday to Friday, the average boardings peak in the afternoon, while Saturday and Sunday do not have a significant peak. Average boardings on Monday in September appear lower compared to other days and months possibly due to the idea that many are returning from vacation or from travelling over the weekend. Average boardings on Wednesday, Thursday, and Friday in November appear lower due to changing weather conditions or seasonal occurrences. 

\newpage

### **Facet Plot 2**

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(capmetro) + 
  geom_point(aes(x=temperature, y=boarding, col=factor(weekend)), size=.75) + 
  facet_wrap(~hour_of_day) +
  labs(title='Boardings Vs. Temperature by Hour of the Day Across Day Type', col='Day Type', x='Temperature (Farenheit)', y='Total Boarding') +
  scale_color_manual(values=c('lightpink','lightblue')) +
  theme_minimal()

```

The faceted scatter plot represents the relationship between temperature (in Farenheit) and total CapMetro boarding, faceted by hour of the day (in 24-hour time) and grouped by day type (weekday and weekend). The scatterplots assess the degree of correlation between temperature and boarding based on hour of the day and day type, visualizing the extent of whether temperature affects total boarding. When hour of the day and weekend status are held constant, temperature does not appear to have a noticeable effect on the number of UT students riding the bus, as the scatterplots remain leveled at similarly the same total boarding as the temperature changes. 

\newpage

## **Problem 4: Wrangling the Billboard Top 100**

### **Part A**

```{r echo=FALSE, warning=FALSE, message=FALSE}

billboard <- billboard %>%
  arrange(desc(weeks_on_chart))

top10songs <- billboard %>%
  distinct(song, .keep_all=TRUE) %>%
  head(n=10)

top10table <- top10songs %>%
  subset(select = -c(X, url, week_id, week_position, song_id, instance, previous_week_position, year, week, peak_position))

kbl(top10table, booktabs=T, col.names=c("Song", "Performer", "Count")) %>%
  kable_styling(latex_options=c("striped", "hold_position"))

```

The table shows the top 10 most popular songs since 1958 up to mid-2021, measured by the total number of weeks that a song spent on the Billboard Top 100. There are three columns, with the song title, performer of the song, and count, or the number of weeks the song appeared in the Billboard Top 100. The songs are ordered in descending order based on the count column to signify the popularity of the songs.

### **Part B**

```{r echo=FALSE, warning=FALSE, message=FALSE}

billboard_sort <- billboard %>% 
  filter(year!=1958 & year!=2021)

unique_songs <- billboard_sort %>%
  group_by(year, song) %>%
  summarize(count=n_distinct(song)) %>%
  group_by(year) %>%
  summarize(total_songs=sum(count))

ggplot(unique_songs) +
  geom_line(aes(x=year, y=total_songs), col='palevioletred3') +
  labs(title='Musical Diversity Over the Years', x='Year', y='Number of Unique Songs in the Billboard Top 100') +
  theme_minimal()

```

The line graph portrays the number of unique songs that appeared in the Billboard Top 100 to measure the degree of 'musical diversity' in the Billboard Top 100 over time. Musical diversity peaks between 1960 and 1970, however decreases overall as time progresses to 2000, where it then increases and reaches a peak again at 2020. Musical diversity was at a overall low near the year 2000. 

### **Part C**

```{r echo=FALSE, warning=FALSE, message=FALSE}

ten_week_hit <- billboard %>%
  filter(weeks_on_chart >= 10)

ten_week_songs <- ten_week_hit %>%
  group_by(performer, song) %>%
  summarize(count=n_distinct(song)) %>%
  group_by(performer) %>%
  summarize(num_of_songs=sum(count))

ten_week_artists <- ten_week_songs %>%
  filter(num_of_songs >= 30) 

ggplot(ten_week_artists) +
  geom_bar(aes(x=performer, y=num_of_songs), col='palevioletred3', fill='lightpink', stat='identity') +
  labs(title='Ten-Week Hits by Performer', x='Performer', y='Number of Ten-Week Hits') +
  coord_flip() +
  theme_minimal()

```

The bar plot consists of 19 artists in the U.S. musical history who had at least 30 songs categorized as 'ten-week hits', or a song that appeared in the Billboard Top 100 for at least ten weeks. The barplot depicts the amount of ten-week hits each performer had. Elton John had the greatest number of ten-week hits, followed by Madonna and Kenny Chesney. 


